---
title: "Import and clean Foreign Assistance"
author: "Laura Hughes"
date: "December 14, 2015"
output: html_document
toc: true
---

## Overview

**In this module, we'll import in the data and start to check that it looks properly.**

The 

### The data


### Functions we'll cover in this module
* library
* read_csv
* head
* kable
* 

## Import in helper functions to make R more powerful
R out of the box (called 'base R') is great. But the really, really powerful thing about R is that they have a community of people helping write other functions to expand R's toolkit and make it better, faster, and more powerful. Before we start playing with data, we'll import some of the most useful functions to import, clean, manipulate, and visualize data.

```{r import fucntions}
# Workhorse libraries for data science: data cleanup, summaries, merging, etc.
library(dplyr) # Filter, create new variables, summarise, ... Basically, anything you can think to do to a dataset
library(tidyr) # Reshape and merge datasets
library(stringr) # String manipulation

# Incredibly powerful plotting library built off of the "Grammer of Graphics"
library(ggplot2)

# Libraries to help import files
library(haven) # Imports in files from Stata, SAS, and SPSS
library(readr) # An advanced form of the base 'read.csv' file with some added functionality.

library(knitr) # Helper function to produce this 
```

## Let's get started and get data into R!

```{r importData}
fileName = '~/GitHub/StataTraining/Data/StataTraining.csv'
spent = read_csv(fileName)

# Print out a table of what the data looks like.
# At the command line, you can type View(spent), which pulls up a table to view the data.


# There's also a function called 'head' which allows you to look at a little bit of the data so you don't have to see the full thing.  It'll show you the first 6 rows of any dataset.

# 'kable' is a function which formats info into a neat table which 
kable(head(spent))
```


```{r fix import}
# Read the help documentation to figure out where the arguments going.
help("read_csv")

spent2 = read_csv(fileName,
                 col_names = TRUE, skip = 2)

# Check that this looks better
kable(head(spent2))
```

Excellent!  That looks a lot better.

---

## Now let's start to take a deeper look at the data.
Let's take another look at our data and make sure that everything is imported correctly.

R has many redundant ways of doing things. Sometimes they're exactly the same 
(with different details under the hood), and sometimes they're slightly different and complementary.

In this case, 'glimpse' and 'summary' provide two quick looks at what a dataset looks like.

```{r look at data}
# 'glimpse' gives you the name of the variable, its type, and the initial values.
glimpse(spent2)

# 'summary' similarly gives you the type of data, but also for numerical data, gives you quick stats on the range of the data, the mean, and the distribution.
summary(spent2)

```

### D'oh! The main column we care about -- amount spent -- isn't a number in our dataset.  Instead, it's a string (a series of alphanumeric characters.)

#### Why doesn't R import the amount as number??
Since amounts are given as things like $5,238.23 in the dataset, they're imported as strings. The dollar sign isn't a number, so R assumes the variable is a series of characters.

#### Before we go any further, we need to fix that.  The easy way first.
The readr package is pretty nifty.  Using it, we can tell the importer function that the Amount column is a number and it'll ignore anything that isn't 0-9 or a decimal point.

```{r fix Amount}
# Simple way-- reimport the data. If you specify the data format for each of the columns, it should take care of the rest.


# You can use a compact string representation where each character represents one column: c = character, i = integer, n = number, d = double, l = logical, D = date, T = date time, t = time, ? = guess, or _/- to skip the column.

# We'll use the 'numeric' class, since it's a sloppy parser that ignores everything except numbers, -, and .

spent3 = read_csv(fileName,
                 col_names = TRUE, skip = 2,
                 col_types = 'iicccccccn')

glimpse(spent3)

```


#### That's a nice in-built function. But you can accomplish the same thing (albeit with more work) using string manipulation.
We'll use some functions within stringr and dplyr to: 
1. Strip out the $ and , and save the value as a new variable.
2. Convert the string to a number and save as a new variable.

##### Part 1: Get rid of the $! (and the ,'s)
```{r remove$}
# More complicated way -- import the data, remove the dollar sign, and convert the strings to numbers.

# We'll work from spent2.

spent4 = spent2 %>% 
  mutate(Amount2 = str_replace(Amount, '$', ''))

# Let's look.
glimpse(spent4)

```

##### Huh.  That's not what we expected.  The $ are still there!  What's going on?
This is actually a pretty typical problem to encounter.  We'll have to track down this problem.  

In this case, it turns out that on the backend, stringr is using what's called [https://en.wikipedia.org/wiki/Regular_expression](Regular Expression Matching) to figure out what to replace.  In regex (as it's known), certain characters like $ are used as special characters to denote special behavior.  In the case of the dollar sign, it's used in conjunction with a ^ to do exact matching.  If you search for 'blue' in a list containing blue and blueberry, you'll get both words.  If instead you search for '^blue$', you'll only get blue (the exact match).

To get around this problem, you stick what's called an escape character in front of the character you want.  This character (\ in normal programming; \\ in R) tells the computer that the next character should be interpreted literally as a dollar sign, not as the special behavior like it would normally.


```{r remove$2}
# How'd I figure that out?  Let's start with a simple test:
str_replace_all(c('$3'), '$', '')

# That doesn't work.  Let's try replacing the 3.
str_replace_all(c('$3'), '3', '')

# Hmm.  That does.  

# !! Moment of clarity-- I remember $ are special in regex expressions.  Normally, you fix this by adding \ before the special character.

str_replace_all(c('$3'), '\$', '')

# Gives error message: Error: '\$' is an unrecognized escape in character string starting "'\$"
# Ugh.  Time to turn to StackExchange --> R uses double escape sequences.  Problem solved!
str_replace_all(c('$3'), '\\$', '')

# Note: the Magrittr operator -- that funky %>% -- is amazing.  It allows you to string together operations from the dplyr package, instead of having to call them sequentially.
spent5 = spent2 %>% 
  mutate(Amount2 = str_replace(Amount, '\\$', ''),
         Amount3 = str_replace(Amount2, ',', ''))

# Let's look.
glimpse(spent5)

```

Ahh, bliss.  But those "" around the number and the 'chr' designation still indicate that the data is a character, not a number.  Time to fix that.

#### Part 2: convert the character to a number.
```{r convert2Num}
spent6 = spent5 %>% 
  mutate(Amount3 = as.numeric(Amount2))

# 
spent = spent2 %>% 
  mutate(amount = str_replace(Amount, '\\(', '-'), 
         amount2 = str_replace_all(amount, '[^\\w-.]', ''),
         amt = as.numeric(amount2)) # tricksy-- replacing anything that isn't a 'word' -- letter or number
```


### Lesson: be careful with your data, and check that what you're doing is correct. Don't save over the original data till you're sure what you have is correct.

In this case, I actually cheated and got lucky. I saw that there were parentheses in the dollar amounts when I was skimming through the data, so I knew to watch out for them.